The Galactic Chemical Evolution of Carbon: Implications for Stellar Nucleosynthesis

### 2025-03-07 carbon_paper/tex/main.tex
c paper updates, fixed solar scale

% Enter the current year, for the copyright statements etc.
\pubyear{2024}

@@

Despite being the most abundant metal in the Universe~\citep[e.g.][]{magg+22}, the nucleosynthetic origin of C is poorly understood. 
There is broad agreement that both high mass and asymptotic giant branch (AGB) stars produce C \citep{jennifer19}, but their relative yields are disputed.
From a Galactic chemical evolution (GCE) perspective, some authors have argued that high-mass stars should dominate~\citep[e.g.][]{prantzos+94, HEK00, romano+20, franchini+20, gustafsson22}, and others have argued that AGB stars should contribute at least half of the C in the universe~\citep[e.g.][]{tinsley79, chiappini+03, mattsson10, KKU11, rybizki+17, KKL20}. 
In this paper, we aim to constrain C production using subgiant stars and multi-zone GCE models \dbstrike{incorporating stellar radial migration}.

@@

Predictions of C yields from stellar evolution and supernova (SN) models are riddled with uncertainties (see, e.g., the reviews by \citealt{romano+10, KL14}).
Some \dbadd{substantial uncertanties} include mass loss rates \citep{sukhbold+16, beasor+2020}, rotational mixing \citep{frischknecht+16, LC18}, nuclear reaction rates \dbadd{CITE}, and convection \citep{chieffi2001, ventura+13}.
Recent work also suggests that small changes in a stars's initial mass of 
 $\sim$$0.01 M_\odot$  have significant effects on the evolution and explosion outcomes of massive stars \citep{bruenn+2023, vartanyan_burrows2023}.
Furthermore, many \add{massive} stars are found in binary systems \dbadd{\url{https://ui.adsabs.harvard.edu/abs/2012Sci...337..444S/abstract} ??}, the effects of which are almost entirely
unexplored in stellar nucleosynthesis models \citet{farmer+21}.
Accurate predictions of metal yields from stellar evolution models remain elusive.

Given the challenge of theoretical yield predictions, recent work aims to empirically characterize elemental production \dbnote{does it make sense to frame this as only recent work? I just don't know but I feel like there are very early examples which at least use observations as a guide for yields, like maybe even B2FH. or garnett1997}
\citet{weinberg+19, weinberg+22} analyzed the abundance ratio trends of high- and low-alpha sequence stars in APOGEE~\citep{apogee17} to infer relative yields of prompt and delayed enrichment sources and the metallicity dependence thereof.
\citet{emily+19, emily+22, emily+24} applied the same methodology to GALAH~\citep{DeSilva2015, Martell2017}.
\citet{rodriguez+21, rodriguez+23} measured the mean Fe yield of core-collapse supernovae (CCSNe) using the radioactive tails of their lightcurves.
Their measurements, provide one of few empirical anchors on the absolute scale of yields \citep{david_fe}.
\dbadd{Additionally the series of papers including \citet{eitner+2020} includes some relevant work.}

 % Folowing many previous works  (e.g. \citealt{DTS78, BF06, prantzos+18, berg+19}; see also the recent reviews by \citealt{romano22} and \citealt{RM21}.) this paper focuses on the evolution of C in GCE models.
\citet{james+24} recently argued that the enrichment history of the Galactic disk can be described as an equilibrium phenomenon.
In equilibrium GCE, trends in abundance ratios closely trace yield ratios (see their section 5.2.2).
\citet{james+23} had previously shown that the correlation between gas-phase N and O abundances (see their Fig. 1) could be described as a consequence of the metallicity dependence of the relative N and O yields.
Under their equilibrium scenario, this argument should extend to any element whose enrichment by stellar populations does not require long timescales ($\gtrsim$few Gyr).
We show in section~\ref{sec:agb} that the characteristic delay times of C production should be $\sim$ 1 Gyr, within this limit. 
\dbnote{This is complicated by the difference between high alpha and low alpha, and 1-2Msun stars do produce carbon...}

We use a sample of subgiant stars from APOGEE \citep{apogee17} as our primary observational constraint.
APOGEE Subgiants provide a large, homogeneous sample of well-measured C abundances unaffected by stellar evolution.
\dbadd{Subgiants' photospheric C abundance closely matches their birth composition.}
\dbstrike{Subgiants have the advantage that their photospheric abundances of C directly reflect their birth C abundances much more closely than other spectral types.} \citep{gilroy89, korn+07, lind+08, souto+18, souto19}.
In other evolutionary phases, the measured abundances are known to be affected by internal processes (see discussion in section 2 below).
Our sample should therefore be unaffected by uncertainties with inferring their birth abundances.

% \dbnote{not sure the observations below need to be  too detailed here? Maybe just state that C/O appears to have a universal shape across environments and direct to that section since our main focus is on APOGEE? and disk chemistry?  }
% \dbstrike{
% From observations of nearby stars and (extragalactic) gas, we know that [C/O] against [O/H] traces a banana shape (see Fig.~\ref{fig:gas_phase} and section~\ref{sec:gas}).
% At the very lowest metallicities ($[{\rm O/H}]\lesssim-2$),%
% %
% observations indicate that [C/Mg] declines with metallicity.
% From $-2 \lesssim {\rm [O/H]}\lesssim -1$, [C/Mg] is roughly constant with metallicity. 
% At higher metallicities ([O/H]~$\gtrsim -1$), [C/Mg] increases with metallicity.
% }
% \dbstrike{We will show in section~\ref{sec:results} below that these correlations in the bulk population abundances alone are quite constraining for the origin of C.}

In Section~\ref{sec:data_selection} we describe the selection for our subgiant observational sample.
In Section~\ref{sec:nucleosynthesis}, we discusses theoretical estimates of \agb\ and \cc\ C yields and present the yield prescriptions we use for our models.
In Section~\ref{sec:vice}, we describe the remaining details of our GCE models.
In Section~\ref{sec:results}, we present our model predictions and discuss their observationally distinguishable predictions. 
In Section~\ref{sec:gas}, we compare our model to observations of halo stars and extragalactic HII regions from the literature. 
We summarize our conclusins in section~\ref{sec:conclusions}.

% \footnotetext{\strike{By metallicity, we mean the (mass) fraction of any element which is not H or He, denoted by $Z$. For the sun, we take $Z_\odot=0.016$.} }

@@

% Subgiants are a useful empirical benchmark because their photospheric abundances should reflect their birth mixture. 
We use subgiant stars as our empirical benchmark.
Subgiants have the advantage that their photospheric C and N abundances should closely approximate their birth mixture.
During main sequence evolution, metals can fall below the convective envelope (\textit{gravitational settling}; e.g. \citealt{turncotte+98}).
When stars evolve off main sequence, these metals are reincorporated into the convective envelope \citep[]{gratton+00, souto19}. 
When a star becomes a red giant, \textit{first dredge up} pollutes its photosphere with core, CNO-processed material
\citep{iben67, KL14}.
\dbnote{Why exactly do we use subgiants? Consistent stellar parameters, subgiant spectra are easier than white dwarfs? only a $\sim 0.05$ dex correction}

@@

Fig.~\ref{fig:subgiants} shows out subgiant sample plotted in the \caah\ and \caafe{} planes.\footnotemark{} In this sample, [C/Mg] increases with [Mg/H], and [C/Mg] decreases with [Mg/Fe] at fixed [Mg/H].

@@

We have also investigated these thrends in other catlogues (not show) in order to understand the systematic uncertainties of our APOGEE subgiant sample.
Namely, we we consider subgiants/dwarfs from GALAH DR4, Gaia-ESO DR5.1, and LAMOST with similar selection criteria. Additionally, we consider APOGEE Giants with first dredge up mixing corrections from \citep{vincenzo+21}.
In general, different surveys follow similar trends in \caah\ and \caafe\ but with systematic offsets at the $\sim{0.1}$ dex level in [C/Mg] given [Mg/H] and slight variations in the slope.
These systematic differences are not a major concern for our conclusions since our models are mostly constrained by the slope of \caah\ and \caafe,  not their normalization (see Fig.~7? and discussion in section \dbadd{5.2}).

@@

A yield quantifies the new production of a chemical species. For a given star, we define the {\it net fractional yield of $X$} as the fraction of a stars initial mass synthesized into new $X$. Specifically, the net fractional yield of element $X$ is
\begin{equation} 
   \y_{\rm X}(M, Z) = 
    \frac{m_{X, \rm ej}- Z_{X,\rm  ini}\,m_{\rm ej}}{m}
\end{equation}
where $Z_X$ is the initial abundance of $X$, $m_{\rm ej}$ is the ejected mass, $m_{X, \rm ej}$ is the ejected mass of $X$, and $m$ is the initial mass of the star.
Note that yields may become negative if more X is transformed into other elements than is produced ($Z_{X,\rm ej} < Z_{X,\rm ini}$).
We also describe yields by their population-averaged quantities, the amount of newly produced X per unit mass of star formation, 
\begin{equation} 
   y_{X}(Z) = 
    \frac{M_X - r\,Z_{X, \rm ini}\,M_{\rm ssp}}{M_{\rm ssp}},
\end{equation}
where $M_X$ is the total expelled mass of element X, $M_{\rm ssp}$ is the birth mass of the single stellar population (SSP\footnote{An SSP is a cluster of stars born at the same time from the same material.}), and $r$ is the return fraction (fraction of the SSP mass returned to the interstellar medium).

Following \citet{james+23}, we integrate yields and our GCE models using the publicly available Versatile Integrator for Chemical Evolution (\VICE, \citealt{JW20}). We use the O, Mg, and Fe yields recommended by \citet{david_fe} based on measurements of the mean CCSN Fe yield \citep{rodriguez+23} and APOGEE chemical abundances.
Our N yield is scaled from \citep{james+23}.
Table~\ref{tab:fiducial_mod} summarizes these choices. 
We assume a \citet{kroupa01} initial mass function.

@@

An Asymptotic Giant Branch (AGB) star is a low mass star undergoing He shell burning, the last nuclear phase (see e.g. \citealt{PR2023}). AGB stars enrich the interstellar medium (ISM) through mass expelled during thermal pulses. 
In this work, we explore four different stellar AGB yield tables from the literature which provide well-sampled grids in mass and metallicity. We refer to the tables as 
\begin{description}
    \item \hypertarget{fruity}{\texttt{FRUITY}}: \citet{cristallo+11, cristallo+15}
    \item \hypertarget{aton}{\texttt{ATON}}: \citet{ventura+13,ventura+14,ventura+18, ventura+20}
    \item \hypertarget{monash}{\texttt{Monash}}: \citet{KL16, karakas+18}
    \item \hypertarget{nugrid}{\texttt{NuGrid}}: \citet{pignatari+16, ritter+18, battino+19, battino+21}
\end{description}
Each AGB yield table reports yields for stellar models on a grid of masses and metallicities as reported in Table~\ref{tab:agb}.
We use the same tables as \citet{james+23}, except that we have swapped \citet{karakas10} for \citet{pignatari+16}. 
For our models to better match observations, we uniformly amplify the yield tables according to
\begin{equation} \label{eq:alpha}
        \ycagb(M, Z) \rightarrow \alpha_{\rm agb}\ \ycagb(M, Z).
\end{equation}
We use the \fruity\ table, with $\alpha_{\rm agb}=\cfactor$, as the fiducial \agb\ yield.
Fig.~\ref{fig:y_agb} compares the C yields for these four AGB models.
Most models agree on general yield trends in mass and metallicity.
C production peak between masses of $\sim$ 2--4 $\Mo$. As metallicity increases, the net yield decreases and the mass of peak C production increases slightly.

In order to integrate AGB yields with \VICE, we linearly interpolate yield tables in both mass and metallicity. Below the smallest mass in a table, $\ycagb$ is linearly extrapolated to 0 at $1\Mo$. The population-averaged AGB star yield is calculated with 
\begin{equation} \label{eq:imf-agb-yield}
    y_{X}^{\rm agb}(Z,\tau) = 
    \frac{1}{A}\int^{u_{\rm AGB}}_{M_\textrm{post-AGB}(\tau)} 
    \y_{\rm X}(M, Z)\ 
    \frac{dN}{dM}\ M\ dM
\end{equation}
are the minimum and maximum mass of star formation, $u_{\rm AGB}=8\Mo$ is the maximum mass of an AGB star, $M_\textrm{post-AGB}(\tau)$ is the mass of stars which complete AGB evolution at age $\tau$, and $dN/dM$ is the initial mass function. $A = \int_l^u M \frac{dN}{dM} dm$ is the IMF normalization coefficient and  $l=0.08\,\Mo$ and $u = 100\,\Mo$. We assume a \citet{larson74} mass-lifetime-relationship and assume that the total lifetime of stars is 1.1 times the MS lifetime.

The left panel of Fig.~\ref{fig:agb-ssp} shows IMF-averaged C yields for each \agb\ model as a function of metallicity.
The IMF-averaged yields differ in normalization and metallicity dependence.  
The normalizations span a factor of \about{2}. (e.g. between $6\times 10^{-4}$ and $12 \times 10^{-4}$ at $Z=0.33\,\Zo$).
Both \nugrid{} and \aton{} predict non-monotonic C yields with $Z$, but it is unclear if this prediction is significant given the uncertainties associated with AGB modelling.
To characterize the metallicity dependence of each table, we fit the yield to a linear model in $\log Z$
\begin{equation}\label{eq:agb_z_approx}
    y_{\rm C}^{\rm AGB}(Z) = \zetao + \zetai \log(Z / \Zo).
\end{equation}
Table \ref{tab:agb} contains the fit parameters $\bar y_{\rm C, 0}^{\rm AGB}$ and $\zagb$ for each AGB table.
The values of $\zagb$ span a factor of $\sim{3}$ between models.
These variations are due to different choices of reaction rates, convection treatments, and mass-loss rates (see discussion below and \JJ).

@@

In AGB stars, {third dredge up} (TDU) and {hot bottom burning} (HBB) are two competing processes driving C evolution.
TDU accompanies thermal pulses, where material from the CO core is mixed with the envelope, increasing surface C abundances 
later released to the ISM \citep{KL14}.  
HBB\ is the activation of the CNO cycle at the bottom of the convective envelope.
TDU increases C yields and HBB converts C into N. \dbstrike{When both processes are active, highly efficient N production ensues (see discussion in e.g. {james+23, ventura+13}).} \dbnote{move to next paragraph?}.
%Because the $^{14}$N proton capture is the slowest component of the CNO cycle, the CNO cycle converts nearly all \C[12] into $^{14}$N \citep{solar-fusion}.

@@

The primary uncertainties in AGB stellar evolution are convection, nuclear reaction rates, and mass loss prescriptions.
\citet{ventura+15} compared the \aton{} and \monash{} models, finding that differences in yields are primarily driven by differences in mass-loss and convection prescriptions. \aton{} predicts strong HBB to set in $\sim{1}\,\Mo$ lower than in \monash{}, leading to much lower C yields.
Moreover, the $^{14}{\rm N}({\rm p},\gamma)$ reaction rate uncertainty causes a factor of 2 difference in predicted C yields \citep{herwig+austin2004, HAL2006} \dbnote{palmerini+11, reaction rates updated since then}.
Most of the physics included in stellar modelling is empirically calibrated. Mass-loss rates, convection, and extra mixing are not sufficiently well understood to be computed from first principles.

Super AGB stars are the class of AGB stars with masses $\sim8$ to $\sim10\Mo$ which burn O into Si and Mg. While we neglect super-AGB stars in the discussion above, these stars are not predicted to be substantial producers of C, similar to high mass AGB stars. From the \citet{doherty+14, doherty+14b} yield table (extending the \monash\ yields), the highest net fractional yield is only -0.002 for super solar metallicity, which would have a negligible impact on the integrated yield). Additionally, as discussed in section~\ref{sec:degeneracies} below, the short lifetimes of high mass AGB and super-AGB stars mean they have negligable impacts on our conclusions and can be grouped with CCSNe yields.

@@

Massive stars form $^{12}$C in their cores through the triple-$\alpha$ reaction \dbadd{(He-burning)} which is later released to the ISM through winds and supernovae.

In \VICE, massive star nucleosynthesis is approximated as occurring instantaneously after a star formation event. Under this approximation, the yield is simply a constant  of proportionality between the metal production rate and the SFR:
\begin{equation}
    \dot{M}_X^{\rm CC} = y_{X}^{\rm CC}\, \dot{M}_\star
\end{equation}
The population-averaged yield, $y_{X}^{\rm CC}$ can be expressed as
\begin{equation}
%y_{X}^{\rm CC} = \frac{ \int_{l_{\rm CC}}^{u} (E(M) M_{X\rm, SN} + w_X - Z_X\,(M-M_{\rm rem}) )\frac{dN}{dM}\ dM }{ \int_{l}^{u} M\,\frac{dN}{dM}\ dM }
y_{X}^{\rm CC} = \int_{l_{\rm CC}}^{u} \left[E(m) m_{X\rm, SN} + w_X - Z_X\,(m-m_{\rm rem})\right]\,\frac{1}{A} \frac{dN}{dm}\ dm
\end{equation}
where $E(m)$ is the fraction of stars of mass $m$ which explode, $M_{X,\rm SN}$ and $w_X$ are the mass of X ejected in a CCSN explosion and through stellar winds respectively, $Z_X$ is the initial abundance of $X$, $dN/dM$ is the IMF, $l_{\rm CC}=8\Mo$ is the initial minimum mass of a CCSN progenitor, $M_{\rm ej}$ is the total mass ejected from the star in winds?. \dbstrike{We integrate yields using \VICE's {\tt vice.yields.ccsne.fractional} function.} \dbnote{is this too detailed?}

@@

\begin{figure*}
    \centering
    \includegraphics{cc_yields.pdf}
    \caption[]{
    \dbnote{y label: integrated CCSNe carbon yield}
    \dbnote{might be interesting to plot every CCSNe yield used in this paper in light grey to show strength of our constraint}
        C yields from high-mass stars.
        \textbf{Left} The IMF-weighted \cc\ yield of C as a function of metallicity.
        \textbf{Right} The \cc\ [C/Mg] abundance ratio, defined in Eq.~\ref{eq:c_mg_cc}. The black line is the derived C yield from section~\ref{sec:mcmc_results}. Yields are shown for tables from 
    \citet[red triangles]{WW95}, \citet[orange square and diamond]{sukhbold+16}, 
    \citet[green stars]{NKT13}, and \citet[blue circles]{LC18}. \citet{sukhbold+16} report yields for different black hole landscapes, while \citet{LC18} provide yields at different rotational velocities.
    In the left panel, the pink line denotes $\Ycagb$ from \fruity{} for comparison. All models include wind yields. 
}
    \label{fig:y_cc}
\end{figure*}

@@

\dbnote{moved this here since next subsection depends on representative sample}
To create representative stellar samples to compare with APOGEE, 
we draw \nsubgiants\ stars from the simulated stellar populations. First, we draw a random zone (weighted by number of subgiants in zone) and then draw a random particle in the zone weighted by SSP mass.
Galactocentric radii are taken from astroNN \citep{leung+bovy19, leung+bovy19b}, however Gaia parallaxes yield a similar distribution. Alternative distributions of stars (including uniform distributions in 7-9kpc) minimally affect median trends.

>>>>

% Enter the current year, for the copyright statements etc.
\pubyear{2025}

@@

Despite being the most abundant metal in the Universe~\citep[e.g.][]{magg+22}, the nucleosynthetic origin of C is poorly understood. 
There is broad agreement that both high mass and asymptotic giant branch (AGB) stars produce C \citep{jennifer19}, but their relative yields are disputed.
From a Galactic chemical evolution (GCE) perspective, some authors have argued that high-mass stars should dominate~\citep[e.g.][]{prantzos+94, HEK00, romano+20, franchini+20, gustafsson22}, and others have argued that AGB stars should contribute at least half of the C in the universe~\citep[e.g.][]{tinsley79, chiappini+03, mattsson10, KKU11, rybizki+17, KKL20}. 
In this paper, we aim to constrain C production using subgiant stars and multi-zone GCE models.% \dbstrike{incorporating stellar radial migration}.

@@

Predictions of C yields from stellar evolution and supernova (SN) models are riddled with uncertainties (see, e.g., the reviews by \citealt{romano+10, KL14}).
Some salient unknowns include mass loss rates \citep{sukhbold+16, beasor+2020}, rotational mixing \citep{frischknecht+16, LC18}, nuclear reaction rates \citep{herwig+05, herwig+austin2004}, and convection \citep{chieffi2001, ventura+13}.
Recent work additionally suggests that small changes in a stars's initial mass of 
 $\sim$$0.01 M_\odot$  have significant effects on the evolution and explosion outcomes of massive stars \citep{bruenn+2023, vartanyan_burrows2023}.
 Furthermore, many stars are found in binary systems \citep{sana+12}, the effects of which are almost entirely
unexplored in stellar nucleosynthesis models \citet{farmer+21}.
Accurate predictions of metal yields from stellar evolution models remain elusive.

Given the challenge of theoretical yield predictions, recent work aims to empirically characterize elemental production. \dbnote{are there older relevant references like garnett1997? or other relevant work (eitner+20?)}
\citet{weinberg+19, weinberg+22} analyzed the abundance ratio trends of high- and low-alpha sequence stars in APOGEE~\citep{apogee17} to infer relative yields of prompt and delayed enrichment sources and their metallicity dependence.
\citet{emily+19, emily+22, emily+24} applied the same methodology to GALAH~\citep{DeSilva2015, Martell2017}.
\citet{rodriguez+21, rodriguez+23} measured the mean Fe yield of core-collapse supernovae (CCSNe) using the radioactive tails of their lightcurves.
Their measurements, provide one of few empirical anchors on the absolute scale of yields \citep{david_fe}.

 % Folowing many previous works  (e.g. \citealt{DTS78, BF06, prantzos+18, berg+19}; see also the recent reviews by \citealt{romano22} and \citealt{RM21}.) this paper focuses on the evolution of C in GCE models.
\citet{james+24} recently argued that the enrichment history of the Galactic disk can be described as an equilibrium phenomenon.
In equilibrium GCE, trends in abundance ratios closely trace yield ratios (see their section 5.2.2).
\citet{james+23} had previously shown that the correlation between gas-phase N and O abundances (see their Fig. 1) could be described as a consequence of the metallicity dependence of the relative N and O yields.
Under their equilibrium scenario, this argument should extend to any element whose enrichment by stellar populations does not require long timescales ($\gtrsim$few Gyr).
We show in section~\ref{sec:agb} that the characteristic delay times of C production should be $\sim$ 1 Gyr, within this limit. 
\dbnote{This is complicated by the difference between high alpha and low alpha, and 1-2Msun stars do produce carbon.... Also, I would like to call out reviews and general works on C at some point.}

We use a sample of subgiant stars from APOGEE \citep{apogee17} as our primary observational constraint.
APOGEE Subgiants provide a large, homogeneous sample of well-measured C abundances unaffected by stellar evolution.
\dbadd{Subgiants' photospheric C abundance closely matches their birth composition.}
\dbstrike{Subgiants have the advantage that their photospheric abundances of C directly reflect their birth C abundances much more closely than other spectral types.} \citep{gilroy89, korn+07, lind+08, souto+18, souto19}.
In other evolutionary phases, the measured abundances are known to be affected by internal processes (see discussion in section~\ref{sec:data_selection} below).
Our sample should therefore be unaffected by uncertainties with inferring their birth abundances.

In section~\ref{sec:data_selection} we describe the selection for our subgiant observational sample.
In section~\ref{sec:nucleosynthesis}, we discusses theoretical estimates of \agb\ and \cc\ C yields and present the yield prescriptions we use for our models.
In section~\ref{sec:vice}, we describe the remaining details of our GCE models.
In section~\ref{sec:results}, we present our model predictions and discuss their observationally distinguishable predictions. 
In section~\ref{sec:gas}, we compare our model to observations of halo stars and extragalactic HII regions from the literature. 
We summarize our conclusins in section~\ref{sec:conclusions}.

% \footnotetext{\strike{By metallicity, we mean the (mass) fraction of any element which is not H or He, denoted by $Z$. For the sun, we take $Z_\odot=0.0176$.} }

@@

% Subgiants are a useful empirical benchmark because their photospheric abundances should reflect their birth mixture. 
We use subgiant stars as our empirical benchmark.
Subgiants have the advantage that their photospheric C and N abundances should closely approximate their birth mixture.
During main sequence evolution, metals can fall below the convective envelope (\textit{gravitational settling}; e.g. \citealt{turncotte+98}).
When stars evolve off main sequence and become subgiants, these metals are reincorporated into the convective envelope \citep[]{gratton+00, souto19}. 
However, once a star becomes a red giant, \textit{first dredge up} pollutes its photosphere with core, CNO-processed material
\citep{iben67, KL14}. 
\dbnote{Is there other justification for subgiants? Consistent stellar parameters, subgiant spectra are easier than white dwarfs? only a $\sim 0.05$ dex correction}

@@

Fig.~\ref{fig:subgiants} shows out subgiant sample plotted in the \caah\ and \caafe{} planes.\footnotemark{} In this sample, [C/Mg] increases with [Mg/H], and [C/Mg] decreases with [Mg/Fe] at fixed [Mg/H]. Our aim is to reproduce these trends.

@@

We have also investigated these thrends in other catlogues (not show) in order to understand the systematic uncertainties of our APOGEE subgiant sample.
Namely, we we consider subgiants/dwarfs from GALAH DR4 \citep{galah_dr4}, Gaia-ESO DR5.1 \citep{gaiaeso_dr5.1}, and LAMOST \citep{lamost} with similar selection criteria. Additionally, we consider APOGEE Giants with first dredge up mixing corrections from \citep{vincenzo+21}.
In general, different surveys follow similar trends in \caah\ and \caafe\ but with systematic offsets at the $\sim{0.1}$ dex level in [C/Mg] given [Mg/H] and slight variations in the slope.
These systematic differences are not a major concern for our conclusions since our models are mostly constrained by the slope of \caah\ and \caafe,  not their normalization (see Fig.~\ref{fig:zeta_f} and discussion in section \dbadd{5.2}).

@@

A yield quantifies the new production of a chemical species. For a given star, we define the {\it stellar yield of $X$} as the fraction of a star's initial mass synthesized into a new element, $X$. Specifically, the stellar yield of $X$ is
\begin{equation} \label{eq:stellar_yield_def}
   \y_{\rm X}(m, Z) := 
    \frac{m_{X, \rm ej}- Z_{X,\rm  ini}\,m_{\rm ej}}{m}
\end{equation}
where $Z_{X,\rm ini}$ is the initial mass abundance of $X$, $m_{\rm ej}$ is the ejected mass, $m_{X, \rm ej}$ is the ejected mass of $X$, and $m$ is the initial mass of the star.
Note that yields may become negative if more X is transformed into other elements than is produced.
We also describe yields by their population-averaged quantities, {\it integrated yields}, the amount of newly produced X per unit mass of star formation. 
The integrated yield of $X$ is
\begin{equation} 
   y_{X}(Z) := 
    \frac{M_X - r\,Z_{X, \rm ini}\,M_{\rm ssp}}{M_{\rm ssp}},
\end{equation}
where $M_X$ is the total expelled mass of element X, $M_{\rm ssp}$ is the birth mass of the single stellar population (SSP)\footnote{An SSP is a cluster of stars born at the same time from the same material.}, and $r$ is the return fraction (fraction of the SSP mass returned to the interstellar medium).

Table~\ref{tab:fiducial_mod} summarizes our yield choices. 
Following \citet{james+23}, we integrate yields and our GCE models using the publicly available Versatile Integrator for Chemical Evolution (\VICE, \citealt{JW20}). We use the O, Mg, and Fe yields recommended by \citet{david_fe} based on measurements of the mean CCSN Fe yield \citep{rodriguez+23} and APOGEE chemical abundances.
Our N yield is scaled from \citep{james+23}.
We assume a \citet{kroupa01} initial mass function (IMF).

@@

An asymptotic giant branch (AGB) star is a low mass star undergoing He shell burning, the last nuclear phase (see e.g. \citealt{PR2023}). AGB stars enrich the interstellar medium (ISM) through mass expelled during thermal pulses.

We explore four different stellar AGB yield tables from the literature which provide well-sampled grids in mass and metallicity. We refer to the tables as 
\begin{description}
    \item \hypertarget{fruity}{\texttt{FRUITY}}: \citet{cristallo+11, cristallo+15}
    \item \hypertarget{aton}{\texttt{ATON}}: \citet{ventura+13,ventura+14,ventura+18, ventura+20}
    \item \hypertarget{monash}{\texttt{Monash}}: \citet{KL16, karakas+18}
    \item \hypertarget{nugrid}{\texttt{NuGrid}}: \citet{pignatari+16, ritter+18, battino+19, battino+21}
\end{description}
Each AGB yield table reports yields for stellar models on a grid of masses and metallicities as reported in Table~\ref{tab:agb}.
We use the same tables as \citet{james+23}, except that we have swapped \citet{karakas10} for \citet{pignatari+16}. (See section~\ref{sec:nugrid_yields} for the implementation and recalculation of these yields.)
For our models to better match observations, we uniformly amplify the yield tables according to
\begin{equation} \label{eq:alpha}
        \ycagb(m, Z) \rightarrow \alpha_{\rm agb}\ \ycagb(m, Z).
\end{equation}
We use the \fruity\ table, with $\alpha_{\rm agb}=\cfactor$, as the fiducial \agb\ yield.

Fig.~\ref{fig:y_agb} compares the C yields for these four AGB models.
Most models agree on general yield trends in mass and metallicity.
C production peak between masses of $\sim$ 2--4 $\Mo$. As metallicity increases, the net yield decreases and the mass of peak C production increases slightly.

In order to integrate AGB yields with \VICE, we linearly interpolate yield tables in both mass and metallicity. Below the smallest mass in a table, $\ycagb$ is linearly extrapolated to 0 at $1\Mo$. The population-averaged AGB star yield is calculated with 
\begin{equation} \label{eq:imf-agb-yield}
    y_{X}^{\rm AGB}(Z,\tau) = 
    \int^{u_{\rm AGB}}_{m_\textrm{post-AGB}(\tau)} 
    \y_{\rm X}(m, Z)\ 
    \frac{1}{A}\frac{dN}{dm}\ m\ dm
\end{equation}
are the minimum and maximum mass of star formation, $u_{\rm AGB}=8\Mo$ is the maximum mass of an AGB star, $m_\textrm{post-AGB}(\tau)$ is the mass of stars which complete AGB evolution at age $\tau$, and $dN/dm$ is the initial mass function (IMF). The IMF normalization is
\begin{equation} \label{eq:imf_normalization}
    A := \int_l^u m \frac{dN}{dm} dm
\end{equation}
    where  $l=0.08\,\Mo$ and $u = 100\,\Mo$ are the assummed minimum and maximum mass of star formation. We assume a \citet{larson74} mass-lifetime-relationship and assume that the total lifetime (through post-AGB) is $1.1\times$ the main sequence lifetime.

The left panel of Fig.~\ref{fig:agb-ssp} shows IMF-averaged C yields for each \agb\ model as a function of metallicity.
The IMF-averaged yields differ in normalization and metallicity dependence.  
The normalizations span a factor of \about{2}. (e.g. between $6\times 10^{-4}$ and $12 \times 10^{-4}$ at $Z=0.33\,\Zo$)
Both \nugrid{} and \aton{} predict non-monotonic C yields with $Z$, but AGB modeling uncertanties may eclipse this prediction's significance. To characterize the metallicity dependence of each table, we fit the yield to a linear model in $\log Z$
\begin{equation}\label{eq:agb_z_approx}
    y_{\rm C}^{\rm AGB}(Z) = \zetao^{\rm AGB} + \zetai^{\rm AGB} \log(Z / \Zo).
\end{equation}
Table \ref{tab:agb} also contains the linear fit parameters for each AGB table.
The values of $\zagb$ span a factor of $\sim{3}$ between models.
These variations are due to different choices of reaction rates, convection treatments, and mass-loss rates (see discussion below and \JJ).

@@

In AGB stars, {third dredge up} (TDU) and {hot bottom burning} (HBB) are two competing processes driving C evolution.
TDU accompanies thermal pulses, where material from the CO core is mixed with the envelope, increasing surface C abundances 
later released to the ISM \citep{KL14}.  
HBB\ is the activation of the CNO cycle at the bottom of the convective envelope.
TDU increases C yields and HBB converts C into N. 
%Because the $^{14}$N proton capture is the slowest component of the CNO cycle, the CNO cycle converts nearly all \C[12] into $^{14}$N \citep{solar-fusion}.

@@

The primary uncertainties in AGB stellar evolution are convection, nuclear reaction rates, and mass loss prescriptions.
\citet{ventura+15} compared the \aton{} and \monash{} models, finding that differences in yields are primarily driven by differences in mass-loss and convection prescriptions. \aton{} predicts strong HBB to set in $\sim{1}\,\Mo$ lower than in \monash{}, leading to much lower C yields.
Moreover, the $^{14}{\rm N}({\rm p},\gamma)$ reaction rate uncertainty causes a factor of 2 difference in predicted C yields \citep{herwig+austin2004, HAL2006}.
Most of the physics included in stellar modelling is empirically calibrated. Mass-loss rates, convection, and extra mixing are not sufficiently well understood to be computed from first principles.

Super AGB stars are the class of AGB stars with masses $\sim8$ to $\sim10\Mo$ which burn O into Si and Mg. While we neglect super-AGB stars in the discussion above, these stars are not predicted to be substantial producers of C, similar to high mass AGB stars. From the \citet{doherty+14, doherty+14b} yield table (extending the \monash\ yields), the highest net fractional yield is only -0.002 for super solar metallicity, which would have a negligible impact on the integrated yield). Additionally, as discussed in section~\ref{sec:degeneracies} below, the short lifetimes of high mass AGB and super-AGB stars mean they have minimally affect our conclusions and can be grouped with CCSNe yields.

@@

Massive stars form $^{12}$C in their cores through the triple-$\alpha$ reaction (He-burning) which is later released to the ISM through winds and supernovae.

We approximate massive star enrichment as occurring instantaneously after a star formation event. Under this approximation, the yield is simply a constant  of proportionality between the metal production rate and the SFR:
\begin{equation}
    \dot{M}_X^{\rm CC} = y_{X}^{\rm CC}\, \dot{M}_\star
\end{equation}
The CCSN integrated yield is given by
\begin{equation}
y_{X}^{\rm CC} = \int_{l_{\rm CC}}^{u} \left[E(m) m_{X\rm, SN} + w_X - Z_X\,(m-m_{\rm rem})\right]\,\frac{1}{A} \frac{dN}{dm}\ dm
\end{equation}
where $E(m)$ is the fraction of stars of mass $m$ which explode, $m_{X,\rm SN}$ and $w_X$ are the mass of $X$ ejected in a CCSN explosion and through stellar winds respectively, $Z_X$ is the initial abundance of $X$, $dN/dm$ is the IMF, $l_{\rm CC}=8\Mo$ is the initial minimum mass of a CCSN progenitor. 
%\dbstrike{We integrate yields using \VICE's {\tt vice.yields.ccsne.fractional} function.} \dbnote{is this too detailed?}

@@

\begin{figure*}
    \centering
    \includegraphics{cc_yields.pdf}
    \caption[]{
    \dbnote{might be interesting to plot every CCSNe yield used in this paper in light grey to show strength of our constraint}
        C yields from high-mass stars.
        \textbf{Left} The IMF-weighted \cc\ yield of C as a function of metallicity.
        \textbf{Right} The \cc\ [C/Mg] abundance ratio, defined in Eq.~\ref{eq:c_mg_cc}. The black line is the derived C yield from section~\ref{sec:mcmc_results}. Yields are shown for tables from 
    \citet[red triangles]{WW95}, \citet[orange square and diamond]{sukhbold+16}, 
    \citet[green stars]{NKT13}, and \citet[blue circles]{LC18}. \citet{sukhbold+16} report yields for different black hole landscapes, while \citet{LC18} provide yields at different rotational velocities.
    In the left panel, the pink line denotes $\Ycagb$ from \fruity{} for comparison. All models include wind yields. 
}
    \label{fig:y_cc}
\end{figure*}

@@

To create representative stellar samples to compare with APOGEE, 
we draw \nsubgiants\ stars from the simulated stellar populations. First, we draw a random zone (weighted by number of subgiants in zone) and then draw a random particle in the zone weighted by SSP mass.
Galactocentric radii are taken from astroNN \citep{leung+bovy19, leung+bovy19b}, however Gaia parallaxes yield a similar distribution. Alternative distributions of stars (including uniform distributions in 7-9kpc) minimally affect median trends.
